---
title: "Random Numbers and Seeds in R"
author: "Jason Bryer"
date: 2020-08-31
draft: false
categories: ["R"]
tags: ["random numbers"]
---

To explore how R generates randome numbers, we will use the `rnorm` function. This function draws a random number from a normal distribution with a mean = 0 and standard deviation = 1 (though these can be changed with the `mean` and `sd` parameters). With `n = 1` we will get two random numbers.

```{r}
rnorm(n = 1)
rnorm(n = 1)
```

Each time you run the command you will get a different number. The `set.seed` function will sets a seed to the random number generator so that each subsequent run will produce the same number.

```{r}
set.seed(2112); rnorm(n = 1)
set.seed(2112); rnorm(n = 1)
```

Setting a different seed results in a different number.

```{r}
set.seed(2113); rnorm(n = 1)
```

**What are seeds?**

Computers are actually bad at random events. However, there are good algorithms that mimic random processes. These algorithms work by starting with some initial value, a seed, and executing a complex algorithm that approximates randomozation. The seed is often set to the current time in miliseconds. To visualize the ramdom process, we will use the `sample` function to randomly select a number between 1 and 100. We will consider the ouput for the first 1,000 seeds.

```{r}
random_numbers <- integer(1000)
for(i in seq_len(length(random_numbers))) {
	set.seed(i)
	random_numbers[i] <- sample(1:100, size = 1)
}
library(ggplot2)
ggplot(data.frame(x = 1:100, y = random_numbers),
	   aes(x = x, y = y)) + 
	geom_point() +
	xlab('Seed') + ylab('Random Number')
```

